---
title: "Final_Project"
author: Hailey Armor, Lucy Karn, and Justin Kruljac
date: 4/29/2025
date-format: long
format:
  pdf:
    code: false
    pdf-engine: lualatex
    geometry: margin=1in
---

```{r setup, include=FALSE, error=FALSE}
# 1. load libraries
      
      library(dplyr)
      library(readr)
      library(readxl)
      library(lubridate)
      library(knitr)
      library(ggplot2)
      library(kableExtra)
      library(tinytex)


# 2. Tidy missed days of school
       
       missed_days <- read_csv("six_or_more_workdays_missed_due_to_illness_injury_or_disability___total___2019_to_2023 (1).csv")

tidy_missed_days <- missed_days %>%
  filter(Subgroup == "18 years and older", Year != 2019) %>%
  select(Year, Missed_Workdays = Estimate)

#3. Tidy COVID data by year 

covid <- read_csv("COVID-19_Daily_Counts_of_Cases__Hospitalizations__and_Deaths (1).csv")
covid <- covid %>%
  mutate(date = mdy(date_of_interest))

##Create year column
covid <- covid %>%
  mutate(Year = year(date))

##Organize total hospitalizations/deaths by year
covid_summary <- covid %>%
  group_by(Year) %>%
  summarize(
    Total_Hospitalizations = sum(HOSPITALIZED_COUNT, na.rm = TRUE),
    Total_Deaths = sum(DEATH_COUNT, na.rm = TRUE)
  ) %>%
    filter(Year >= 2020) 
# 4. Combine the two tidy tables
combined_data <- left_join(tidy_missed_days, covid_summary, by = "Year")
# 5. Create graph of COVID Hospitalization and Missed Days of School
graph <- ggplot(combined_data, aes(x = Year)) +
  geom_col(aes(y = Total_Hospitalizations), fill = "lightblue", width = 0.6) +
  geom_line(aes(y = Missed_Workdays * 10000), color = "red", linewidth = 1.2) +
  scale_y_continuous(
    name = "Total Hospitalizations",
    sec.axis = sec_axis(~./10000, name = "% Missing 6+ Work/school Day")
  ) +
  labs(
    title = "COVID Hospitalizations vs Missed Workdays (2020-2023)",
    x = "Year"

  ) + 
  theme_minimal()
      
# 6. Calculate the correlation
correlation <- cor(combined_data$Total_Hospitalizations, combined_data$Missed_Workdays, use = "complete.obs")

cat("Correlation coeffecient for relationship between COVID hospitalizations and missed workdays:", round(correlation, 3))
```

## Overview

The goal of this repo is to provide the readers with a statistical insight of the data provided from the National Center of Health Statistics (NCHS). We also would like to compare this data to the number of COVID cases in NYC in the same years. From these two data sets we will calculate a difference in proportions and determine if it is statistically significant.

## Provenance

One data set is from the National Center of Health Statistics, it contains data ranging from 2019-2023 regarding the the frequency of missed work due to illness, injury, or disability over the given time period. The other data set is from NYC Dept. of Health COVID Report and it contains cases of COVID starting in 2020 and records them daily. Each case in this data set is a day, where as each case in the first data set is a year.

We would like to point out a possible association between an increase the number of work days missed and the post-COVID era. The purpose of this doc is to create a correlation table and data visualization for these two data sets to see and prove that there is a correlation between sick days and COVID cases.

## Fair and Care Principles

Our main data set of number of people under 18 taking a sick day from school adheres to the FAIR principles because it is accessible, findable, and reusable. This data set is available to the public and uses language that is easily understood. It can also be reused to observe smaller groups such as race, location, age, and health insurance.

Our second data set of COVID cases adheres more to the CARE principles. It focuses more on data collected for collective benefit and for the government to use to track COVID. While this data set doesn't exactly capture or reflect directly on other cultures and worldviews, it can be used to draw conclusions about COVID that relate to how COVID is moving worldwide. This data set was collected ethically and is going to be for future use.

## Data Analysis

```{r}
#| label: tbl-tidy-missed-days
#| tbl-cap: "Percentage of Adults Missing 6+ Workdays Due to Illness or Injury by Year"
#| echo: false
#| error: false
#| 
kbl(tidy_missed_days, format = "latex", booktabs = TRUE) %>%
  kable_styling(latex_options = c("scale_down"), full_width = FALSE, position = "center")
```

In @tbl-tidy-missed-days, the percent of adults missing 6 or more days of work each year are recorded. This data has been recorded from 2019 to 2023, but this table gets rid of the data from 2019 because it is not relevant to our research question and other data. This table shows us that there is an increase overall since 2020, however the peak year was 2022 and not 2023. This suggests that the there was an increase during 2020, 2021, and 2022, but there is now a decrease according to the 2023 data.

```{r}
#| label: tbl-covid-summary
#| tbl-cap: "COVID Hospitalizations and Deaths by Year (2020–2023)"
#| echo: false
#| error: false

kbl(covid_summary, format = "latex", booktabs = TRUE) %>%
  kable_styling(latex_options = c("scale_down"), full_width = FALSE, position = "center")
```

In @tbl-covid-summary, the number of COVID hospitalizations and deaths in NYC is recorded per year. We are using these number to simulate the trend of COVID cases across the nation. This data ranges from 2020 to 2025, but we will only be using data from 2020-2023 to match our other data set. This table shows us that the highest number of hospitalizations and deaths occurred in 2020 and the lowest occurred in 2024 (excluding 2025 because it has not been a full year yet). The table demonstrates how each year the number of deaths and hospitalizations decreases, signifying that as time increases the rate of COVID deaths and hospitalizations will decrease.

```{r}
#| label: tbl-combined-data
#| tbl-cap: "Combined COVID Hospitalizations and Missed Workdays by Year"
#| echo: false
#| error: false
kbl(combined_data, format = "latex", booktabs = TRUE) %>%
  kable_styling(latex_options = c("scale_down"), full_width = FALSE, position = "center")
```

In @tbl-combined-data, the two previous data sets are merged, so that they each only have data from 2020-2023. This table shows the missed workdays for adults vs. COVID hospitalizations and deaths each year. The side by side numbers allow us to see that the COVID numbers decrease the entire time and the missed workdays increase and then decrease.

```{r}
#| label: fig-graph
#| fig-cap: "COVID Hospitalizations vs Missed Workdays"
#| fig-align: "center"
#| echo: false
#| error: false
print(graph)
```

In @fig-graph, the number of COVID hospitalizations is represented by blue bars at the bottom of the graph and the percent of adults missing 6+ workdays is represented by a red line on the graph. This graph doesn't show any immediate clear explanation or similar patterns, but after looking closer there is a huge decrease from 2022 to 2023 for COVID hospitalization and this is also the only time when percent of adults missing 6+ days of work decreases. There could be correlation, such that when COVID hospitalization are high, the percent of adults missing 6+ days of work increases, but when hospitalizations decreases significantly than the percent of adults missing 6+ workdays will decrease.

```{r}
#| label: correlation
#| echo: false
#| error: false
cat("Correlation coeffecient for relationship between COVID hospitalizations and missed workdays:", round(correlation, 3))
```
A correlation coefficient of -0.234 signifies a weak negative correlation. This statistic does not support our initial assumption that an increase in COVID hospitalizations in NYC is positively correlated with an increase in proportion of the population (18+) missing six or more workdays due to injury/illness/disability.

## Conclusion

Our data analysis (calculating a correlation coefficient) does not yield a result that supports our initial assumptions. However, this is a surface level analysis. There are several other questions that can be asked. There are also several other variables including confounding variables that can be accounted for in a further analysis. It is also important to account for the fact that the NCHS data only pertains to those 18+, while age is not a variable in the NYC dataset. It is also important to note that we are comparing quantity COVID hospitalizations, not cases, in NYC, which may ultimately impact results. Our correlation coefficient (r = -02.43) suggests a weak negative correlation between our data. However, despite what value is computed for our correlation coefficient in this analysis, causation cannot be proven as this is an observational study and only pertains to a certain population. Ultimately the conclusion yields that more research must be done, often this is the conclusion that many analysts face in all fields before reaching a statistically significant result.


## Data Sources and Acknowledgements

Data from:

NCHS Data Query System. Six or more workdays missed due to illness, injury, or disability [Internet]. Hyattsville (MD): National Center for Health Statistics; c2024 [cited 2025 May 4]. Available from: https://www.cdc.gov/nchs/dqs

U.S. Department of Health and Human Services. COVID-19 daily counts of cases, hospitalizations, and deaths [Internet]. Washington (DC): Data.gov; [cited 2025 May 4]. Available from: https://catalog.data.gov/dataset/covid-19-daily-counts-of-cases-hospitalizations-and-deaths

## Authors

Hailey Armor hpa52552\@psu.edu 
Lucy Karn lek5494\@psu.edu 
Justin Kruljac jck5770\@psu.edu

## Code Appendix

```{r codeAppendix, ref.label=c("setup", "load-missed-days","load-covid-summary","load-combined-data", "load-final-graph" ), echo=TRUE, eval=FALSE, error=FALSE}
     
```
